---
author:
- name: Sangeeta Bhatia
date: "`r Sys.Date()`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
spacing: double
bibliography: 
biblio-style: apsr
endnote: no
output: html_document
---

```{r setup, eval = TRUE}
library(tidyverse)
library(ggthemes)
library(viridis)
library(gganimate)
```

# Data wrangling

```{r}
global_mortality <- here::here("week_3/data", "global_mortality.xlsx") %>%
  readxl::read_xlsx(.)
```

Clean up the names for easy access but save the names for pretty 
plotting.

```{r}
names(global_mortality) <- str_trim(
  str_remove_all(names(global_mortality), "[[:punct:]]")
)
causes <- colnames(global_mortality)[ -c(1, 2, 3)]



```

Wide to tall so that we can group by countries or causes of death etc.

```{r}

mortality_tall <- tidyr::gather(
  data = global_mortality,
  key = "cause",
  value = "percent",
  -c(country, countrycode, year),
  factor_key = TRUE
  )

```
Sum by year for each country.

```{r}
mortality_country <- group_by(mortality_tall, country, cause) %>%
  summarise(percent = sum(percent, na.rm = TRUE))
```
Rows where country code is NA are actually groups of countries rather
than a single country. The groups are overlapping (e.g., countries are
grouped according to geographic region as well as SDI). We can explore
each group separately.



```{r}
regions <- c(
  "Andean Latin America",
  "Central Latin America",
  "Tropical Latin America",
  "Latin America and Caribbean",
  "Southern Latin America",
  "North America",
  "Caribbean",
  "Australasia",
  "Oceania",
  "East Asia",
  "South Asia",
  "Central Asia",
  "Southeast Asia",
  "Eastern Europe",
  "Central Europe",
  "Western Europe",
  "North Africa and Middle East",
  "Eastern Sub-Saharan Africa",
  "Central Sub-Saharan Africa",
  "Western Sub-Saharan Africa",
  "Southern Sub-Saharan Africa",
  "Sub-Saharan Africa"
)

country_groups <- filter(mortality_tall, country %in% regions) %>%
    droplevels()

```

Re-arrange country groups geographically.

```{r}
country_groups$country <- factor(
  country_groups$country,
  levels = regions
)
```
## Leading causes of death in each geographical region

```{r, fig.show = "animate", interval = 2}

p <- ggplot(country_groups, aes(country, cause, frame = year)) +
  geom_tile(aes(fill = percent)) +
  scale_fill_viridis(name = "% Population") +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("County Group") + ylab("Cause") +
  theme(legend.title = element_blank())

gganimate(p)

```

## Leading causes of death by SDI

```{r, fig.show = "animate", interval = 2}
sdi <- c(
  "High SDI",
  "High-middle SDI",
  "Middle SDI",
  "Low-middle SDI",
  "Low SDI"
)

sdi_groups <- filter(mortality_tall, country %in% sdi) %>%
  droplevels()

sdi_groups$country <- factor(
  sdi_groups$country,
  levels = sdi
)

p2 <- ggplot(sdi_groups, aes(country, cause, frame = year)) +
  geom_tile(aes(fill = percent)) +
  scale_fill_viridis(name = "% Population") +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("") + ylab("Cause")

gganimate(p2)

```

## Leading causes of death in the UK

```{r}
uk <- c("England", "Northern Ireland", "Scotland", "Wales")
uk_groups <- filter(mortality_tall, country %in% uk) %>%
  droplevels()

```
Let us look at the top 10 causes rather than the whole data set.
```{r}
top_10 <- group_by(uk_groups, country, year) %>%
    top_n(n = 10, wt = percent)  %>%
    droplevels() %>%
    pull(cause) %>% unique

uk_groups <- filter(uk_groups, cause %in% top_10) %>% droplevels()

```

```{r fig.show = "animate", interval = 2}
uk_groups$country <- factor(
  uk_groups$country,
  levels = uk
)

p3 <- ggplot(uk_groups, aes(percent, cause, col = country, frame = year)) +
      geom_point(aes(size = percent)) + theme_tufte()

gganimate(p3)

```
